TTC delays: when they happen and how to be punctual

### 2024-09-23 paper/paper.qmd
update cleaned data

# Load the data
subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))

# Convert the Time column to a proper time format
subwayData$Time <- hm(subwayData$Time)  # Parse 'Time' as hours and minutes
busData$Time <- hm(busData$Time)

# Extract the hour from the Time column
subwayData$Hour <- hour(subwayData$Time)
busData$Hour <- hour(busData$Time)

@@

# Plot 1: Average delays by hour of day (combined subway and bus)
ggplot(hourly_avg_combined_long, aes(x = factor(Hour), y = Mean_Delay, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Delays by Hour of Day (Subway vs Bus)", 
       x = "Hour of Day", 
       y = "Average Delay (Minutes)", 
       fill = "Type") +
  scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
  theme_minimal() +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, face = "bold"))
```
Two main takeaways from the figure: one, that the magnitude of delays is generally the same with each other, and two, that the average
delay time for buses is around 4-5 times more than that of subways. This confirms the intuition that subways generally are a more time-reliable way to get somewhere than buses, and we will therefore focus on subways from hereonwards in our quest for the most reliable transit plan.

While we don't know the total frequency of subways and therefore cannot calculate an "average delay per ride" statistic, what we can do is sum up the delay at each hour of day to still obtain an estimator for how much delay you're likely to encounter in a trip during each respective hour, as shown in @fig-times-total.

```{r}
#| label: fig-times-total
#| fig-cap: total delay times
#| echo: false

@@

```{r}
#| label: fig-times-adjusted
#| fig-cap: Delay time adjusted for the approximate subway load
#| echo: false
adjust_subway_delays <- function(data) {
  data %>%
    mutate(Adjusted_Delay = ifelse(Hour %in% 6:9 | Hour %in% 15:19, Min.Delay / 2, Min.Delay))
}

>>>>

# Load the cleaned data
subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))

@@

# Plot 1: Average delays by hour of day (combined subway and bus)
ggplot(hourly_avg_combined_long, aes(x = factor(Hour), y = Mean_Delay, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Delays by Hour of Day (Subway vs Bus)", 
       x = "Hour of Day", 
       y = "Average Delay (Minutes)", 
       fill = "Type") +
  scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
  theme_minimal() +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, face = "bold"))
```
Two main takeaways from the figure: one, that the average magnitude of delays is generally the same over the day, whether it be for subways or buses, and two, that the average delay time for buses is around 4-5 times more than that of subways. This confirms the intuition that subways generally are a more time-reliable way to get somewhere than buses, and we will therefore focus on subways from hereonwards in our quest for the most reliable transit plan.

While we don't know the total frequency of subways and therefore cannot calculate an "average delay per ride" statistic, what we can do is sum up the delay at each hour of day to still obtain an estimator for how much delay you're likely to encounter in a trip during each respective hour, as shown in @fig-times-total. 
```{r}
#| label: fig-times-total
#| fig-cap: total delay times
#| echo: false

# Load the cleaned data
subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))

@@

```{r}
#| label: fig-times-adjusted
#| fig-cap: Delay time adjusted for the approximate subway load
#| echo: false

# Load the cleaned data
subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))

# Function to adjust subway delays
adjust_subway_delays <- function(data) {
  data %>%
    mutate(Adjusted_Delay = ifelse(Hour %in% 6:9 | Hour %in% 15:19, Min.Delay / 2, Min.Delay))
}
